Staying Hydrated: The Molecular Journey of Gaseous Sulfur Dioxide to a Water Surface

### 2012-03-29 resubmission/Article.tex
Finalized changes and revisions for resubmission. Only need to make a TOC graphic.

In this study we use ab initio quantum molecular dynamics (MD) techniques to model and simulate the hydrating structures that form around a surface-bound \suldiox~on a water cluster surface. We simulate a dynamic water surface of a small cluster, complete with all the extended hydrogen-bonding interactions that capture the variability of the \suldiox~hydrate structures, and the behavior of water's surface molecules. The quantum MD technique described herein allows more accurate and realistic simulation than our previous classical MD.\cite{Shamay2011} It is also superior to small cluster DFT studies because it does not assume geometry optimized configurations, and the extended interactions of a larger water cluster are incorporated. In our previous classical MD study we determined net orientational behavior of \suldiox~binding to a water surface, and the orientation of the waters as they respond to the presence of an adsorbing gas. Understanding the orientational behavior of molecules in the aqueous interfacial region during adsorption was a necessary first step to understanding the specific details of gas-binding and surface behavior.

%Unlike small cluster studies that use DFT calculations on geometry optimized structures to find likely geometries in a static vacuum setting, The quantum MD technique allows us to simulate the process much more realistically, and with greater accuracy than our previous classical MD work could allow.\cite{Shamay2011} This computational study greatly enhances the picture developed in our previous work on \suldiox.

@@

On-the-fly ab initio molecular dynamics simulations were performed with the QUICKSTEP package (Car-Parrinello method of MD), which is an implementation of the Gaussian plane wave method using the Kohn-Sham formulation of density functional theory (DFT).\cite{VandeVondele2005} The Kohn-Sham orbitals are expanded using a linear combination of atom-centered Gaussian-type orbital functions. The electronic charge density was described using an auxiliary basis set of plane waves. Energies and forces from on-the-fly simulation sampling of the Born-Oppenheimer surface were calculated for each MD step using the Gaussian TZV2P basis set with a cutoff of 280 Ry, the exchange-correlation functional of Becke, Lee, Yang, and Parr (B3LYP),\cite{LEE1988} and the atomic pseudo-potentials of the Goedecker, Teter, and Hutter type.\cite{Goedecker1996} A simulation timestep of 0.5 fs was used, with a Nos\'{e}-Hoover thermostat set at 210K and 300K for the ``cold'' and ``hot'' simulations, respectively, using a coupling constant of 1000 fs. %These computational parameters were verified to yield a reasonable description of bulk room temperature water when simulating a neat-water system.

Initially, 20 equilibrated boxes of side-lengths 8.9545\angs, with 24 randomly packed water molecules were used. Ten of the boxes were used for each of the cold and hot simulations. A sulfur dioxide molecule was randomly placed onto the surface within 2.5\angs~of a water molecule centrally located above the waters in the z-axis. A copy of the initial system cubes were then expanded along one axis (z-axis) to 30\angs. The system energy was minimized through a geometry optimization. Subsequently, the system was equilibrated for 1 ps in canonical ensemble (NVT) conditions. Periodic boundaries were set on the two short axes to form an infinite slab. The equilibrated systems were then simulated for a further 20 ps in the microcanonical ensemble (NVE), with trajectory snapshots recorded every 0.5 fs. The initial 1 ps equilibration trajectory was not included in the final analysis. This simulation process resulted in 40,000 time steps of system trajectory for analysis in each of the hot and cold replicas of the system, for a total of 400,000 timesteps at each temperature.

@@

The bonding coordination of each \suldiox~was determined at each timestep of the simulations. Figure \ref{fig:bonding-coordinations} shows the distribution of bonding coordinations of the surface \suldiox, as a percentage of all bonding coordinations encountered for both the cold (blue) and hot (red) trajectories. A first visual inspection reveals several trends. At both temperatures the \suldiox~spends much of the time unbound without any bonding to the surface waters. The cluster of the three most populated coordinations is comprised of the ``S'', ``SO'', and ``SOO'' bonding types. This grouping represents the coordinations with a single bond through the sulfur, and up to two interactions through the \suldiox-oxygens. In this coordination group the ``SO'' is the most populous, but in no way dominates the bonding types as the difference in populations is less than 2\% and 5\% within the cold and hot simulations, respectively. The second and third most populated bonding coordinations (excluding the unbound state) are the ``S'' and ``SOO'', however their distributions differ between temperatures. In cold simulations the ``S'' and ``SOO'' coordinations occur nearly equally (18.8\% and 18.1\%, respectively). The hotter temperature simulation shifts the distribution such that the ``S'' occurs 4.6\% less frequently than in the cold, and the ``SOO'' occurs 8.6\% less often. The distribution of bonded coordinations (those excluding the ``unbound'' state) in the hot temperature has a clear first and second most frequent coordination: ``SO'' and ``S'', respectively.

@@

% Baer's assymetric oxygen binding
Baer et al. performed this coordination analysis for their single-temperature study, but discriminated between \suldiox~binding through the two different oxygens. They concluded that there is asymmetric hydrogen bonding through the \suldiox-oxygens, with one oxygen binding more often than the other. This is supported by the findings here where the double oxygen coordinations (e.g. ``OO'', ``SOO'', etc.) represent a lower percentage of the coordinations than the single oxygen counterparts (e.g. ``O'', ``SO'', etc.). Furthermore, a triple-oxygen coordination (e.g. ``OOO'', ``SOOO'', etc.) is very rarely encountered. Three \suldiox-oxygen bonds only form if both \suldiox-oxygens are interacting with water hydrogens. Our finding that triple-oxygen coordinations rarely form, and that double-oxygen coordination are less frequent than the single-oxygen counterparts, complements Baer et al's conclusion about the asymmetry in the oxygen interactions.

@@

%On the cold water surface, the ``S'' and ``SO'' are more populated than for the hot system. The relative decrease of these coordinations are matched in the hot surface by an increase of the ``SOO'' configuration. This speaks to a dramatic difference in the surface behavior of \suldiox~at the two temperatures. The cold \suldiox~spends nearly equal time in the ``S'' and ``SOO'' coordinations, but nearly 20\% more time in the ``SO''. Thus, the addition or removal of a bond through the \suldiox-oxygen to a neighboring coordination (e.g. addition of an oxygen bond from ``SO'' to ``SOO'', or removal of the bond from ``SO'' to ``S'') is equally probable, as long as the sulfur interaction with the water oxygen does not break.

@@

As expected for the more populated coordinations, there are generally more transitions between larger nodes in Figure \ref{fig:coordination-transitions} than transitions to less populated coordinations. Insights to the bonding process are made clearer from these graphs. In the cold system graph of Figure \ref{fig:coordination-transitions}, we find that the majority of transitions are between the unbound and ``O'' nodes, and between the single-sulfur coordinations (e.g. ``S-SO'', ``SO-SOO'', etc.) The number of transitions within these groups of bonding coordinations is an order of magnitude larger than any other transition. This indicates that while the \suldiox~is bound in any of the most populated bonding coordinations, it is actively binding and unbinding the oxygens to form the other coordinations, typically those with a single sulfur interaction. Clearly, the \suldiox-sulfur bond to water oxygens is stronger than its oxygen bonds to water-hydrogens. The \suldiox-oxygens are actively and constantly forming and breaking bonds to neighboring waters. In the graph of the hot system in Figure \ref{fig:coordination-transitions} (right), the transition frequencies follow the same trend as in the cold system, increasing with adjacent node size. %One very surprising result is in the transition from ``SOO-SOOO''. This transition frequency does not follow from the adjacent node sizes, as the ``SOOO'' node represents less than 5\% of the bonding coordinations. This is indicative of a very rapid cycle of forming and breaking of bonds to the second \suldiox-oxygen. As noted earlier, it is likely that ``SOO'' coordinated \suldiox, asymmetrically binding twice through a single oxygen, is being pulled further into the water interface. It is likely more surrounded by waters, and in the hot system it can more easily form a brief third hydrogen bond to a water through the second \suldiox-oxygen. Because the triple-oxygen coordination is not as favorable, it quickly breaks the bond and the \suldiox~returns to the ``SOO'' coordination.

@@

We know that the \suldiox~bound to a water surface is most likely in the ``SO'' bonding coordination, and is also often taking part in some type of cyclic structure. With this in mind, it becomes interesting to ask: how long does a cyclic hydrate form before breaking to an acyclic hydrate structure? To answer the question of cyclic lifespan, a method was devised to define the lifetime of a cycle. For each MD trajectory, coordinate data was analyzed to determine if a cyclic hydrate structure was formed as described earlier in this manuscript, and the lifespan data was collected as was done for the bonding coordinations. %A timeline was then produced where each timestep was given a value of 1 or 0 depending on whether a \suldiox~hydrate bonding cycle was present or not, respectively. This resulted in a time-function, $C(t)$, similar in nature to a time-varying digital signal.

%In an electro-mechanical system, a mechanical switch often outputs a noisy signal, full of transients or ``signal bounce'' before settling to a final value. This will appear as a rapid on-off cycling of the signal, and the problem is one of great concern in signal processing. Many mechanical, electrical, and software solutions have been devised to suppress the transients, or ``debounce'' the signal. Analogously, the hydrate bonding cycles formed in the simulated system often undergo a period of time during formation, or before dissolution, where the $C(t)$ function bounces before settling into a final value. The bounce in the function manifests itself in our statistics as a series of rapid switches between cycle presence and cycle absence. Physically, the bond lengths within the cycle (or the cycle-to-be) are fluctuating back and forth across our bond-length criteria. It is thus an artifact of our algorithmic determination of the presence or absence of a bond. Removing this artifact will allow us to examine the longer-time bonding behavior and this is accomplished as follows.

%We wish to find a function based on $C(t)$ that eliminates the brief oscillations during transitions. The resulting function, $f(t)$, will only contain information of whether a hydrate bonding cycle is present or absent, and none of the noisy oscillations of the transitions between the two states. $f(t)$ may then be used to calculate statistics about the cycle lifespans. A representative portion of a cycle time-function, $C(t)$, is plotted for one of the simulated trajectories shown as the dashed black line in Figure \ref{fig:debouncing}. At the far left of the plot the function is in the ``no cycle'' state indicating that a bonding cycle has not been detected, and then switches ``on'' as a cycle is found later in time. The cyclic structure is very dynamic, constantly moving and distorting, so any of the bonds forming the cyclic bonding structure are liable to break and reform quickly. This bouncing between states is manifested in Figure \ref{fig:debouncing} as a series of sharp spikes in the $C(t)$ function lasting less than 10 fs each. $C(t)$ was smoothed using a moving gaussian window function with a 10 fs width, having the effect of disregarding cycle breaks or formations of less than 20 fs duration. The resulting smoothed function, $C_s(t)$, was then cutoff with the following criteria:

%\[
%  f(t) = \left\{ 
%  \begin{array}{l l}
%    0 & \quad C_s(t) < 0.2\\
%    1 & \quad C_s(t) \geq 0.2\\
%  \end{array} \right.
%\]

%where $f(t)$ is the debounced time-function that represents the lifespans of cycles. Figure \ref{fig:debouncing} shows the original time-function of cycle formation and breaking, $C(t)$ (dashed black), the smoothed function, $C_s(t)$ (red), and the final debounced function, $f(t)$ (green).

%\begin{figure}[h!]
%	\begin{center}
%		\includegraphics[scale=1.0]{cycle-debouncing-small.png}
%		\caption{The dynamic environment of the water surface causes rapid breaking and formation of bonds to the \suldiox, and to waters involved in the cyclic hydrating structures. A debouncing procedure was developed (described in the text) to account for the very brief ($<$ 20 fs) breaks to the cyclic structure. A time function, $C(t)$ (dashed black), shows if the bonding distance criteria are met for all members of the cyclic hydrate structure. A gaussian smoothing window was then used to produce $C_s(t)$ (red), which was then used with a cutoff criteria to give the debounced cycle formation time function, $f(t)$ (green).}
%		\label{fig:debouncing}
%	\end{center}
%\end{figure}

@@

We know from the transition frequency plots of Figure \ref{fig:coordination-transitions} that the bonding coordinations are switching frequently. The \suldiox~is likely forming and breaking bonds with waters external to the cyclic hydrate structures (i.e. not directly involved in the bonds of the cycle). For the longer-lived cycles, the external bonding to the \suldiox~may have little effect on the cyclic hydrate waters. However, any time the \suldiox~switches into an unbound, sulfur-only, or oxygen-only coordination (i.e. ``S'', ``SS'', ``O'', etc), the cyclic structure is necessarily broken. Because the majority of cyclic structures last only briefly ($<1$ ps), the active switching of \suldiox~bonding coordinations appears to break the cyclic structure. Figure \ref{fig:bonding-coordinations} shows that the sum of coordinations that necessarily break cyclic structures account for approximately 51\% and 67\% of the bonding coordinations in the cold and hot system, respectively. Consequently, the distribution of Figure \ref{fig:cyclic-breakdown} also indicates that there are more cycles formed in the hot system (nearly 2.5\% above the cold). Of the most encountered bonding coordinations the hot system has a much higher proportion of unbound coordinations, and a relative decrease of the ``SOO'' coordination. The increased temperature appears to cause the \suldiox~to bond in a way that is slightly more conducive to the formation of cyclic structures. %This may account for the slightly increased populations of the longer cycle lifespans in the hot system (greater than 4 ps in some cases) compared to the cold system in the cycle lifespan distribution of Figure \ref{fig:cycle-lifespans}.

>>>>

In this study we use ab initio quantum molecular dynamics (MD) techniques to model and simulate the hydrating structures that form around a surface-bound \suldiox~on a water cluster surface. We simulate a dynamic water surface of a small cluster, complete with all the extended hydrogen-bonding interactions that capture the variability of the \suldiox~hydrate structures, and the behavior of water's surface molecules. The quantum MD technique described herein allows more accurate and realistic simulation than our previous classical MD.\cite{Shamay2011} It is also superior to static DFT studies because it does not assume geometry optimized configurations, and the extended interactions of a larger water cluster are incorporated. In our previous classical MD study we determined net orientational behavior of \suldiox~binding to a water surface, and the orientation of the waters as they respond to the presence of an adsorbing gas. Understanding the orientational behavior of molecules in the aqueous interfacial region during adsorption was a necessary first step to understanding the specific details of gas-binding and surface behavior.

@@

On-the-fly ab initio molecular dynamics simulations were performed with the QUICKSTEP package (Car-Parrinello method of MD), which is an implementation of the Gaussian plane wave method using the Kohn-Sham formulation of density functional theory (DFT).\cite{VandeVondele2005} The Kohn-Sham orbitals are expanded using a linear combination of atom-centered Gaussian-type orbital functions. The electronic charge density was described using an auxiliary basis set of plane waves. Energies and forces from on-the-fly simulation sampling of the Born-Oppenheimer surface were calculated for each MD step using the Gaussian TZV2P basis set with a cutoff of 280 Ry, the exchange-correlation functional of Becke, Lee, Yang, and Parr (B3LYP),\cite{LEE1988} and the atomic pseudo-potentials of the Goedecker, Teter, and Hutter type.\cite{Goedecker1996} A simulation timestep of 0.5 fs was used, with a Nos\'{e}-Hoover thermostat set at 210K and 300K for the ``cold'' and ``hot'' simulations, respectively, using a coupling constant of 1000 fs.

Initially, 20 equilibrated cubes of side-lengths 8.9545\angs, with 24 randomly packed water molecules were used. Ten of the boxes were used for each of the cold and hot simulations. A sulfur dioxide molecule was randomly placed onto the surface within 2.5\angs~of a water molecule centrally located above the waters in the z-axis. A copy of the initial system cubes were then expanded along one axis (z-axis) to 30\angs. The system energy was minimized through a geometry optimization. Subsequently, the system was equilibrated for 1 ps in canonical ensemble (NVT) conditions. Periodic boundaries were set on the two short axes to form an infinite slab. The equilibrated systems were then simulated for a further 20 ps in the microcanonical ensemble (NVE), with trajectory snapshots recorded every 0.5 fs. The initial 1 ps equilibration trajectory was not included in the final analysis. This simulation process resulted in 40,000 time steps of system trajectory for analysis in each of the hot and cold replicas of the system, for a total of 400,000 timesteps at each temperature.

@@

The bonding coordination of each \suldiox~was determined at each timestep of the simulations. Figure \ref{fig:bonding-coordinations} shows the distribution of bonding coordinations of the surface \suldiox, as a percentage of all bonding coordinations encountered for both the cold (blue) and hot (red) trajectories. A first visual inspection reveals several trends. At both temperatures the \suldiox~spends much of the time unbound without any bonding to the surface waters. The cluster of the three most populated bonded coordinations is comprised of the ``S'', ``SO'', and ``SOO'' bonding types. This grouping represents the coordinations with a single bond through the sulfur, and up to two interactions through the \suldiox-oxygens. In this coordination group the ``SO'' is the most populous, but in no way dominates the bonding types as the difference in populations is less than 2\% and 5\% within the cold and hot simulations, respectively. The second and third most populated bonding coordinations (excluding the unbound state) are the ``S'' and ``SOO'', however their distributions differ between temperatures. In cold simulations the ``S'' and ``SOO'' coordinations occur nearly equally (18.8\% and 18.1\%, respectively). The hotter temperature simulation shifts the distribution such that the ``S'' occurs 4.6\% less frequently than in the cold, and the ``SOO'' occurs 8.6\% less often. The distribution of bonded coordinations (those excluding the ``unbound'' state) in the hot temperature has a clear first and second most frequent coordination: ``SO'' and ``S'', respectively.

@@

Baer et al. performed this coordination analysis for their single-temperature study, but discriminated between \suldiox~binding through the two different oxygens. They concluded that there is asymmetric hydrogen bonding through the \suldiox-oxygens, with one oxygen binding more often than the other. This is supported by the findings here where the double oxygen coordinations (e.g. ``OO'', ``SOO'', etc.) represent a lower percentage of the coordinations than the single oxygen counterparts (e.g. ``O'', ``SO'', etc.). Furthermore, a triple-oxygen coordination (e.g. ``OOO'', ``SOOO'', etc.) is very rarely encountered. Three \suldiox-oxygen bonds only form if both \suldiox-oxygens are interacting with water hydrogens. Our finding that triple-oxygen coordinations rarely form, and that double-oxygen coordination are less frequent than the single-oxygen counterparts, complements Baer et al's conclusion about the asymmetry in the oxygen interactions.

@@

@@

As expected for the more populated coordinations, there are generally more transitions between larger nodes in Figure \ref{fig:coordination-transitions} than transitions to less populated coordinations. Insights to the bonding process are made clearer from these graphs. In the cold system graph of Figure \ref{fig:coordination-transitions}, we find that the majority of transitions are between the unbound and ``O'' nodes, and between the single-sulfur coordinations (e.g. ``S-SO'', ``SO-SOO'', etc.) The number of transitions within these groups of bonding coordinations is an order of magnitude larger than any other transition. This indicates that while the \suldiox~is bound in any of the most populated bonding coordinations, it is actively binding and unbinding the oxygens to form the other coordinations, typically those with a single sulfur interaction. Clearly, the \suldiox-sulfur bond to water oxygens is stronger than its oxygen bonds to water-hydrogens. The \suldiox-oxygens are actively and constantly forming and breaking bonds to neighboring waters. In the graph of the hot system in Figure \ref{fig:coordination-transitions} (right), the transition frequencies follow the same trend as in the cold system, increasing with adjacent node size.

@@

We know that the \suldiox~bound to a water surface is most likely in the ``SO'' bonding coordination, and is also often taking part in some type of cyclic structure. With this in mind, it becomes interesting to ask: how long does a cyclic hydrate form before breaking to an acyclic hydrate structure? To answer the question of cyclic lifespan, a method was devised to define the lifetime of a cycle. For each MD trajectory, coordinate data was analyzed to determine if a cyclic hydrate structure was formed as described earlier in this manuscript, and the lifespan data was collected as was done for the bonding coordinations.

@@

We know from the transition frequency plots of Figure \ref{fig:coordination-transitions} that the bonding coordinations are switching frequently. The \suldiox~is likely forming and breaking bonds with waters external to the cyclic hydrate structures (i.e. not directly involved in the bonds of the cycle). For the longer-lived cycles, the external bonding to the \suldiox~may have little effect on the cyclic hydrate waters. However, any time the \suldiox~switches into an unbound, sulfur-only, or oxygen-only coordination (i.e. ``S'', ``SS'', ``O'', etc), the cyclic structure is necessarily broken. Because the majority of cyclic structures last only briefly ($<1$ ps), the active switching of \suldiox~bonding coordinations appears to break the cyclic structure. Figure \ref{fig:bonding-coordinations} shows that the sum of coordinations that necessarily break cyclic structures account for approximately 51\% and 67\% of the bonding coordinations in the cold and hot system, respectively. Consequently, the distribution of Figure \ref{fig:cyclic-breakdown} also indicates that there are more cycles formed in the hot system (nearly 2.5\% above the cold). Of the most encountered bonding coordinations the hot system has a much higher proportion of unbound coordinations, and a relative decrease of the ``SOO'' coordination. The increased temperature appears to cause the \suldiox~to bond in a way that is slightly more conducive to the formation of cyclic structures.
